---
title: "Check-in 1: Instructions"
author: "Peter Gao"
date: "2023-08-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## A. Running your first R code

Your first check-in has two parts. First, you'll try running and reading some basic R code.

1. On the computer you are using, create a new folder titled "MATH167R" or something similar.

2. Within this folder, create a sub-folder named "checkins" or something similar.

3. Download "01_first-R-code.R" and save it in your checkins folder.

4. Open "01_first-R-code.R" in RStudio. It should look something like the code below.

```{r}

```

5. 

5. Answer the two questions 

## B. Compiling your first R Markdown report

Now, we'll use a tool called R Markdown to generate a report that combines code, output, and writing. R Markdown reports are commonly used to share data analysis in a clear and reproducible way.

1. Download 01_first-report.Rmd to your checkins folder and open it in RStudio.

2. 

```{r cars}
summary(cars)
```

To receive full points on the first check-in, make sure:

1. You answered the questions on your first R code here.
2. You submitted your first R Markdown report (in .html form) here.
