---
title: "Check-in 3"
author: "YOUR NAME HERE"
date: "2023-09-11"
output: html_document
---

You should edit this .Rmd using RStudio, then click *Render* in the menu bar of the Source window (above the text of this .Rmd). Remember, you must submit *both* your .Rmd and the compiled .html in order to receive full credit!

### Collaborators

INSERT NAMES OF ANY COLLABORATORS

## A. Weather Forecast Data

**1.**  For this lab, we'll be using the `flights` data. You can load this data using the following code:

    ```{r}
    library(nycflights13)
    data(flights)
    ```

**2.** Use the `mutate()` function to turn `origin` into a factor.

    ```{r}
    # YOUR CODE HERE
    ```

**3.** Compute the mean `arr_delay` for each origin airport. Which airport has the longest delays on average? You can drop missing rows.

    ```{r}
    # YOUR CODE HERE
    ```
    
    <!--# YOUR ANSWER/EXPLANATION HERE -->
    
    
**4.** Use the `mutate()` function to turn `month` into a factor.

    ```{r}
    # YOUR CODE HERE
    ```
    
    <!--# YOUR ANSWER/EXPLANATION HERE -->
    
**5.** Compute the mean `arr_delay` by month. What do you notice?

    ```{r}
    # YOUR CODE HERE
    ```
    
    <!--# YOUR ANSWER/EXPLANATION HERE -->
    
**6.** Compute the mean `arr_delay` by month AND airport. What do you notice?
    